Efficient pedigree recording for fast population genetics simulation

### 2018-01-14 forwards_paper.tex
in progress

%% Progressive memory reduction as a function of T
\citet{padhukasahasram2008exploring} obtained impressive speedups relative to contemporary simulators
by retaining only 8 generations of history.
This begs the question: how fast does the information in the pedigree become irrelevant?
Suppose instead that we have run a simulation of $N$ individuals for $T$ generations,
begun with no prior history.
One way to answer this question is to ask:
How many of the resulting $2NT$ edges are required after simplification?
As above, the expected number of edges is bounded by $N-2$
plus four times the sum of the lengths of all marginal trees,
because new edges appear with most recombinations that occur on any marginal tree.
Therefore, we need to compute the mean total length of all branches in a coalescent tree
up until time $T$ (or the common ancestor, whichever comes first).
Since $k$ lineages coalesce at rate $\binom{k}{2}/N$,
the expected length of time for which a marginal tree has $k$ branches is
$N/(k(k-1)) = N(1/(k-1) - 1/k)$ generations.
Since there are $k$ branches for this time,
this period is expected to contribute $N/(k-1)$ generations to the total tree length.
By summing over $n < k \le N$,
the $N$ tips of a tree are expected to descend from only $n$ lineages
around $N(1/(n-1) - 1/(N-1))$ generations ago.
Inverting this relationship between time and number of roots
implies that a marginal tree cut $T$ units of time ago
is expected to have around $n(T)$ roots, where $n(T) \approx 1 + N/(T+2)$.
The total tree length over this time is
$N \sum_{k=n(T)+1}^{N-1} 1/k$, which
% is approximately
% \begin{align*}
%     N \log\left( \frac{N/2}{1 + N/(T+2)} \right) .
% \end{align*}
leads to an upper bound on the number of edges of
\begin{align}
    \label{eqn:edge_bound}
    N \left( 1 + 4 \log\left( \frac{NT}{2(T + N)} \right)\right) .
\end{align}
These calculations hold up quite well in practice,
as shown in Figure~\ref{fig:simplify_complexity}.
% by comparison to
% the number of edges actually required for 50 independent Wright--Fisher simulations,
% as a function of time (obtained from Algorithm~\algref{W} with $s=1$)
% is shown in Figure~\ref{fig:simplify_complexity}, compared to this prediction.

>>>>

%% Progressive memory reduction as a function of T
\citet{padhukasahasram2008exploring} obtained impressive speedups relative to contemporary simulators
by retaining only 8 generations of history.
This begs the question: how fast does the information in the pedigree become irrelevant?
Suppose instead that we have run a simulation of $N$ individuals for $T$ generations,
begun with no prior history.
One way to answer this question is to ask:
How many of the resulting $2NT$ edges are required after simplification?
As above, the expected number of edges is bounded by $N-2$
plus four times the sum of the lengths of all marginal trees,
because new edges appear with most recombinations that occur on any marginal tree.
Therefore, we need to compute the mean total length of all branches in a coalescent tree
up until time $T$ (or the common ancestor, whichever comes first).
Since $k$ lineages coalesce at rate $\binom{k}{2}/N$,
the expected length of time for which the coalescent process has $k$ lineages is
$2N/(k(k-1)) = 2N(1/(k-1) - 1/k)$ generations.
Since this is the time period over which the tree has $k$ branches,
this period is expected to contribute $2N/(k-1)$ generations to the total tree length.
By summing over $n < k \le N$,
the $N$ tips of a tree are expected to descend from only $n$ lineages
around $2N(1/n - 1/N)$ generations ago.
Inverting this relationship between time and number of roots
implies that a marginal tree cut $T$ units of time ago
% since 2N(1/n - 1/N) = 2(N/n-1) = T implies
%    N/n = (T+2)/2 and so n = 2N/(T+2)
is expected to have around $n(T)$ roots, where $n(T) \approx 2N/(T+2)$.
The total tree length over this time is
$2N \sum_{k=N}^{n(T)+1} 1/(k-1)$, which
% is approximately
% \begin{align*}
%     2 N \left\{ \log(N-1) - \log(n(T)) \} = 2 N \log( ((T+2)(N-1))/2N ) \approx 2 N \log( (T+2)/2 )
% \end{align*}
leads to an upper bound on the number of edges of
\begin{align}
    \label{eqn:edge_bound}
    N \left( 1 + 4 \log\left( \frac{NT}{2(T + N)} \right)\right) .
\end{align}
These calculations hold up quite well in practice,
as shown in Figure~\ref{fig:simplify_complexity}.
% by comparison to
% the number of edges actually required for 50 independent Wright--Fisher simulations,
% as a function of time (obtained from Algorithm~\algref{W} with $s=1$)
% is shown in Figure~\ref{fig:simplify_complexity}, compared to this prediction.
